A Hybrid Neutronics Method for Time-Dependent Control Rod Modeling in Molten Salt Reactors

### 2024-06-11 hybrid.tex
sn terms

The $S_N$ method applies a collocation method for the angular discretization by solving for the
angular flux at $N_d$ discrete directions $\hat{\Omega}_d$ with weights $w_d$. Numerous options
exist for the choice of the angular quadrature set \left{\hat{\Omega}_d,w_d | d=1,\dots,N_d\right}.
For this work, we chose the commonly used level-symmetric quadrature set
\cite{wang_rattlesnake_2018} for ease of implementation, particularly when applying reflective
boundary conditions due to rotational symmetry about each of the three Cartesian axes.

The discrete ordinates ($S_N$) method for solving the multigroup neutron transport equation
(Eq.\ \ref{eq:mg-bte}) discretizes the continuous angular directional phase space into a finite
number of discrete angular intervals (ordinates). The set of distinct direction variables
$\hat{\Omega}_d$ representing the discrete ordinates is typically chosen in conjunction with a
compatible quadrature set to replace integrals across continuous solid angles
$\hat{\Omega}$ with quadrature approximations. The multigroup, discrete ordinates ($S_N$
approximation) form of the 3-D time-dependent neutron transport equation with Legendre polynomial
expansion of the scattering cross section and spherical harmonic expansion of the flux in the
scattering source term is:
%
\begin{gather}
  \frac{\partial}{\partial t} \left(\frac{\Psi_{g,d}(\vec{r},t)}{v_g}\right) +
  \hat{\Omega}_d\cdot\nabla\Psi_{g,d}(\vec{r},t) + \Sigma_{t,g}(\vec{r},t)\Psi_{g,d}(\vec{r},t) =
  \nonumber \\
  \sum^G_{g'=1}\sum^L_{l=0} \frac{2l+1}{4\pi}\Sigma^{g'\rightarrow g}_{s,l}(\vec{r},t)\sum^l_{m=-l}
  \phi_{g',l,m}(\vec{r},t)Y_{l,m}(\hat{\Omega}_d) + \frac{\chi_g(\vec{r},t)}{4\pi}\sum^G_{g'=1}
  \nu\Sigma_{f,g'}(\vec{r},t) \phi_g(\vec{r},t) + S_{g,d}(\vec{r},t)
  \label{eq:3d-sn}
  \shortintertext{where}
  \begin{align*}
    \Psi_{g,d}(\vec{r},t) &= \Psi_g(\vec{r},\hat{\Omega}_d,t) \\
%    g &= \mbox{ neutron energy group index, ranging from 1 to G,} \nonumber \\
%    \Sigma_{t,g}(x) &= \mbox{ macroscopic total cross section of neutrons in group $g$,} \nonumber \\
%    G &= \mbox{ total number of energy groups,} \nonumber \\
%    n &= \mbox{ discrete ordinate index,} \nonumber \\
    N &= \mbox{ total number of discrete ordinates,} \\
    l &= \mbox{ Legendre expansion order,} \\
    L &= \mbox{ highest Legendre expansion order,} \\
    \Sigma^{g'\rightarrow g}_{s,l}(x) &= \mbox{ $l$-th Legendre moment of the macroscopic
scattering} \nonumber \\
    &\ \ \ \ \mbox{ cross section from group $g'$ to $g$,} \\
    \phi_{g',l,m} &= \mbox{ spherical harmonic expansion of the angular flux of degree $l$ and
    order $m$,} \\
    Y_{l,m} &= \mbox{ real basis functions of spherical harmonics of degree $l$ and order $m$.}
%    \chi_g &= \mbox{ fission neutron spectrum in group $g$,} \nonumber \\
%    \nu &= \mbox{ neutrons produced per fission reaction,} \nonumber \\
%    \Sigma_{f,g}(x) &= \mbox{ macroscopic fission cross section of neutrons in group $g$,}
%    \nonumber \\
%    k &= \mbox{ neutron multiplication factor.} \nonumber
%    \phi_{g'}(x) &= \mbox{ scalar neutron flux in group $g$,} \nonumber \\
%    S_g(x,\mu_n) &= \mbox{ external neutron source in group $g$.} \nonumber
  \end{align*}
\end{gather}
%
We can retrieve the neutron scalar flux and current by calculating the 0th and 1st Legendre
moments, respectively, as follows:
%
\begin{align}
  \phi_g(x) &= \frac{1}{2} \int^1_{-1} d\mu\ \Psi_g(x,\mu) = \frac{1}{2} \sum^N_{n=1} w_n
\Psi_g(x,\mu_n)
  \shortintertext{and}
  J_g(x) &= \frac{1}{2} \int^1_{-1} d\mu\ \mu\Psi_g(x,\mu) = \frac{1}{2} \sum^N_{n=1} w_n
\mu_n \Psi_g(x,\mu_n)
\end{align}

The 1-D form of the multigroup $k$-eigenvalue neutron diffusion equations (Eq.\ \ref{eq:mg-diff})
is:
%
\begin{align}
  -\frac{d}{dx} D_g(x) \frac{d}{dx} \phi_g(x) + \Sigma_{t,g}(x) \phi_g(x) &= \sum^G_{g'=1}\left[
  \Sigma_s^{g'\rightarrow g}(x)\phi_{g'}(x) + \chi_g\frac{\nu\Sigma_{f,g'}(x)}{k}
  \phi_{g'}(x)\right] + S_g(x)
  \label{eq:1d-diff}
  \shortintertext{where}
    D_g(x) &= \frac{1}{3 \Sigma_{tr}(x)} \nonumber \\
           &= \mbox{ isotropic neutron diffusion coefficient for group }g. \nonumber
\end{align}

>>>>

The $S_N$ method applies a collocation method for the angular discretization by solving for the
angular flux along $N_d$ discrete directions $\hat{\Omega}_d$ with weights $w_d$. Numerous options
exist for the choice of the angular quadrature set
$\{\hat{\Omega}_d,w_d | d=1,\dots,N_d\}$.
For this work, we chose the commonly used level-symmetric quadrature set
\cite{wang_rattlesnake_2018} for ease of implementation, particularly when applying reflective
boundary conditions due to 90-degree rotational symmetry about each of the three Cartesian axes.

We use the level-symmetric quadrature sets to approximate integrals of the angular flux over the
angular domain
%
\begin{gather}
  \phi_g = \int_\mathcal{S} \Psi_{g,d}\hat{\Omega} \approx \sum^{N_d}_{d=1}w_d
  \Psi_{g,d}), \label{eq:0th-mmt}
  \shortintertext{where}
  \Psi_{g,d} = \Psi_g(\hat{\Omega}_d). \nonumber
\end{gather}
%
We calculate the scalar flux $\phi_g$ by taking the zeroth moment with respect to $\hat{\Omega}$ as
shown in Eq.\ \ref{eq:0th-mmt}. We can also calculate higher moments
%
\begin{gather}
  \phi_{g,l,m} = \int_\mathcal{S} Y_{l,m}(\hat{\Omega})\Psi_g(\hat{\Omega})d\hat{\Omega} \approx
  \sum^{N_d}_{d=1}w_d Y_{l,m}(\hat{\Omega}_d)\Psi_{g}(\hat{\Omega}_d) \label{eq:nth-mmt}
  \shortintertext{where}
  \begin{align*}
    Y_{l,m}(\mu,\omega) &=
    \begin{cases}
      \sqrt{2}C^{|m|}_l P^{|m|}_l \cos(|m|\omega), & \mbox{ for } 0<m\leq l \\
      C^0_l P^0_l(\mu), & \mbox{ for } m=0, 0\leq l < \infty \\
      \sqrt{2} C^{|m|}_l P^{|m|}_l(\mu) \sin(|m|\omega), & -l \leq m < 0
    \end{cases}, \\
    \mu &= \mbox{cosine of the polar angle,} \\
    \omega &= \mbox{the azimuthal angle,} \\
    C^{|m|}_l &= \sqrt{\frac{\left(l-|m|\right)!}{\left(l+|m|\right)!}}, \\
    P^m_l &= \mbox{associated Legendre polynomial of degree $l$ and order $m$.}
  \end{align*}
\end{gather}
%
for anisotropic scattering treatment.

\subsubsection{Final Forms of the \gls{SAAF} $S_N$ Equations}

Finally, we present the expanded forms of each term in the \gls{SAAF} formulation of the multigroup
$S_N$ neutron transport equations. With the angular discretization scheme laid out, we define the
inner products
%
\begin{gather}
  \left(\bm{a},\bm{b}\right)_\mathcal{D} \equiv \sum^G_{g=1} \int_\mathcal{D}d\vec{r}
  \sum_{i\in E_e}a_{g,i}b_i(\vec{r})\sum_{j\in E_e}b_{g,j}b_j(\vec{r}), \\
  \left(\bm{a},\bm{b}\right)_{\partial\mathcal{D}} \equiv \sum^G_{g=1}
  \sum_{s\in\partial\mathcal{D}}\int_s d\vec{r}\sum_{i\in E_s}a_{g,i}b_i(\vec{r})\sum_{j\in E_s}
  b_{g,j}b_j(\vec{r}),
\end{gather}
%
involving integrals over the spatial domain to explicitly show the $S_N$ angular discretization.

Time derivative term:
%
\begin{gather}
  \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,
  \frac{\partial}{\partial t}\left(\frac{\bm{\Psi}}{\bm{v}}\right)\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d}+\tau_g\hat{\Omega}_d\cdot\Psi^*_{g,d},
  \frac{\Psi_{g,d}}{v_g}\right)_\mathcal{D}
\end{gather}
%
Streaming term:
%
\begin{gather}
  \left(\mathbb{L}_1\bm{\Psi}^*,
  \left(\bm{\tau}\mathbb{L}_1-\mathbb{I}+\bm{\tau}\mathbb{L}_2\right)\bm{\Psi}\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\hat{\Omega}_d\cdot\Psi^*_{g,d},\tau_g\hat{\Omega}\cdot
  \nabla\Psi_{g,d}-(1-\tau_g\Sigma_{t,g})\Psi_{g,d}\right)_\mathcal{D}
\end{gather}
%
Collision term:
%
\begin{gather}
  \left(\mathbb{L}_2\bm{\Psi}^*,\bm{\Psi}\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d},\Sigma_{t,g}\Psi_{g,d}\right)_\mathcal{D}
\end{gather}
%
Scattering term:
%
\begin{gather}
  \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,\mathbb{S}\bm{\Psi}\right) =
  \sum^G_{g=1}\sum^{N_d}_{d=1}w_d\left(\Psi^*_{g,d}+\tau_g\hat{\Omega}_d\cdot\Psi^*_{g,d},
    \sum^G_{g'=1}\sum^L_{l=0}\Sigma^{g'\rightarrow g}_{s,l}\sum^l_{m=-l}
